% \setlength{\voffset}{-1.75in}
\chapter{Acknowledgements}
\markboth{\MakeUppercase{Acknowledgements}}{\MakeUppercase{Acknowledgements}}

Many people have, in one way or another, contributed for me to be in a position to do the research I describe in this thesis.
From the servitor to the cleaning staff to numerous fellow scientists that during millennia pursued Truth, and on whose mighty shoulders I have stood.
Andrew Rambaut is one of these rare people that possess both a powerful intellect and a kind soul.
For his generosity in dedicating countless hours to discussing various aspects of my research and helping me grow as scientist, I'm thankful.
Special thanks are in order to Gytis Dudas, Darren Obbard, Jarrod Hadfield, Richard Whittet, Tom Booker, Trevor Bedford, Matthew Hall, Guy Baele, Philippe Lemey and Marc Suchard for stimulating discussions.
Thanks to Lisa for single-handedly handing in this thesis.

I could acknowledge a bunch of my plonker friends by name, but this would inevitably lead to me forgetting someone(s) and getting in trouble.
So I won't.
You know who you are.
My loving wife's support was vital over these arduous years.
To you, \textit{Tatu}, I'm very very grateful.
The unwavering support of my parents and siblings not only during my PhD but throughout my whole life helped me realise my boyhood dream: becoming a professional scientist.

The quotes at the beginning of each chapter are loosely related with the topic of the chapter in question, but don't waste your time trying to make a connection if one is not obvious to you; chances are it only makes sense to myself.
